Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-18 aistats.tex
Add proof for gaussian MLE 1d

Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
\begin{align}
	\phi(z) := \half (z - 1 - \log(z))
\end{align}
illustrated in Figure~\ref{fig:phi}.
\begin{figure}[ht]
	\centering
	\includegraphics[width=.4\textwidth]{phi.pdf}
	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it admits another upper bound (in grey) despite being poorly approximated by quadratics.}
	\label{fig:phi}
\end{figure}

For MAP, we get an upper bound thanks to
\begin{align}
	\label{eq:log_bound}
	\phi(z) \leq \phi(z) + \phi(\inv{z}) =  \half (z + \inv{z}) - 1 \; .
\end{align}

>>>>

Let us recall the form of the divergence with different parametrizations
\begin{align}
	\KL( \sigma_*^2 ; \sigma_n^2 ) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \nat_n; \nat_* ) 
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
\begin{align}
	\phi(z) := \half (z - 1 - \log(z))
\end{align}
illustrated in Figure~\ref{fig:phi}.
\begin{figure}[ht]
	\centering
	\includegraphics[width=.4\textwidth]{phi.pdf}
	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it admits another upper bound (in grey) despite being poorly approximated by quadratics.}
	\label{fig:phi}
\end{figure}

To get an upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we focused on the symmetrized Bregman instead
\aligns{
	\label{eq:log_bound}
	\phi(z) \leq \phi(z) + \phi(\inv{z}) 
	&=  \half \paren{z -1  + \inv{z} -1} 
	= \frac{(z-1)^2}{z}
}
\aligns{
	\bregmanconj( \mu_*; \mu_n) 
	&\leq \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) \\ 
	=: \cS_{\conj} (\mu_*, \mu_n )  
	&= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 }
}

\subsection{Proof of MLE Upper Bound}

	\begin{theorem}[MLE Upper Bound]
	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$ and otherwise
	\begin{align}
		 \expect{\bregmanconj( \mu_*; \mu_n) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
	\end{align}
	\end{theorem}

	\begin{proof}
	The ratio to the optimum  $\frac{n \mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$. Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation 
	\begin{align}
		\expect{n \frac{\mu_*}{n \mu_n}  - 1} 
		= \expect{\frac{n}{\chi^2(n)} - 1} =
		\begin{cases}
			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
			+\infty \  \text{ otherwise. }
		\end{cases}
	\end{align}
	There is also a closed form solution for the expected logarithm of a Chi-squared, because a Chi-squared is a special case of Gamma distributions, whose sufficient statistic is $(X,\log(X))$. For non-central chi-squares, one could also look at \citep{pav2015moments}.
	\begin{align}
		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
	\end{align}
	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
	Consequently the suboptimality of the MLE has a closed form solution
	\begin{align}
	\expect{\bregmanconj( \mu_*; \mu_n) }
	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
	& =
	\begin{cases}
		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
			+\infty \  \text{ otherwise. }
	\end{cases}
	\end{align}
	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
	\begin{align}
		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
	\end{align}
	  For $n\geq 3$,
	\begin{align}
		\inv{n-2} - \inv{n}
		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
		\leq \inv{n-2} - \inv{2 n} \\
		\iff
			\frac{2}{n(n-2)}
			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
			\leq \inv{2n} +\frac{2}{n(n-2)}
	\end{align}
	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
	\end{proof}
